---
title: "Stage M2 -  Relating patterns in woody stem respiration with bark plant functional
  traits in Cerrado"
author: "Edouard DISTIN CARVALHO"
date: "2025-02-25"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


# Effect of fire on Stem respiration 

```{r directory, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/U051-S857/OneDrive/Documents/Fac/Cours/UM/Stage M2/Dados/Stem Respiration")
```

## Data 

```{r charging and cleaning data, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(MASS)
library(lme4)
library(glmmTMB)
library(DHARMa)
library(emmeans)
library(ade4)
library(tidyr)


# Stem respiration data 
stem <- read.csv("C:/Users/U051-S857/OneDrive/Documents/Fac/Cours/UM/Stage M2/Dados/Stem Respiration/stem_efflux_output.csv")  
stem <- stem %>% 
  filter(!is.na(flux_umolm2sec))
stem <- stem[,-c(1,10)]
stem <- stem %>% 
  rename(CO2_efflux = flux_umolm2sec)
stem$date <- as.Date(stem$date, format = "%Y-%m-%d")
stem <- stem %>%
  filter(CO2_efflux <= 5,CO2_efflux >= -1, year(date) == 2023)
stem$sub_plot <- paste(stem$plot_code, "-", stem$sub_plot)

stem <- stem %>%
  mutate(fire_regime = plot_code)

stem$fire_regime <- factor(stem$fire_regime, 
                                    levels = c("ESA-04", "ESA-05", "ESA-06",
                                               "ESA-07", "ESA-08", "ESA-09"),
                            labels = c("Unburned", "Biennial", "Unburned", 
                                      "Triennial", "Unburned", "Annual"))

color <- c( "Biennial" = "green4" , "Triennial" = "blue3", "Annual" = "red", "Unburned" = "black")

```

## Plot 

We conserve only the 2023 data to effect of fire after long time burned.The 2024 data was very different than 2023 because of the drought of the year wich affect the data.

```{r plot of 2023 }


ggplot(data = stem, aes(x = date, y = CO2_efflux, color = fire_regime)) +
  geom_point() +   
  labs(title = "Variation of stem respiration of tree species along the time",
       y = "CO2 efflux (µmol/m2_sec)", x = "Month") + 
  theme_classic() +
  scale_color_manual(values = color)

stem <- stem %>%
  mutate(time = lubridate::hm(datetime))

ggplot(stem, aes(x = time, y = CO2_efflux, color = fire_regime)) +
  geom_point() +
  labs(title = "Variation of stem respiration of trees species in the day",
    x = "Time (HH:MM)",
    y = "Stem respiration (µmol/m2_sec)") +
  scale_x_time(breaks = breaks_width("1 hour"),labels = label_time("%H:%M")) +
  theme_light() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_color_manual(values = color)

ggplot(data = stem, aes(x = air_temp_c, y = CO2_efflux, color = fire_regime)) +
  geom_point() + labs(title = "Variation of stem respiration of trees species across the temperature",
    y = "C02 efflux (µmol/m2_sec)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_color_manual(values = color)

```


## Modelisation 

### Fire effect on stem respiration 

#### Fire Statut : burned or unburned 
variable to explain : CO2_efflux 
Repsonse variable : plot_code - fire_statut - species
Random effect : tree_tag

```{r Condition of GLMM }

ggplot(stem, aes(x = CO2_efflux)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") +
  labs(title = "Distribution of CO2 eflux",
       x = "CO2 efflux",
       y = "Frequence") +
  theme_classic()

```

```{r GLMM fire statut sp}
stem$fire_statut <- ifelse(stem$plot_code %in% 
                             c("ESA-05", "ESA-07", "ESA-09"), "Burned", "Unburned")

mod_fire <- glmmTMB(CO2_efflux ~ fire_statut + (1|species) + (1|tree_tag),
                       data = stem, family = gaussian(link = "identity"))

# Checking the residuals of the models 
plot(simulateResiduals(fittedModel = mod_fire))

residuals_fire <- residuals(mod_fire)

ggplot(data = data.frame(residuals_fire), aes(x = residuals_fire)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Residuals histogram", x = "Residuals values", y = "Frequency") +
  theme_minimal()

summary(mod_fire)

#Test Tukey emmeans fire statut sp}
emmeans_fire <- emmeans(mod_fire, ~ fire_statut)

result_fire <- as.data.frame(pairs(emmeans_fire, adjust = "tukey"))
result_fire

estimate_fire <- summary(emmeans_fire)
estimate_fire

ggplot(estimate_fire, aes(x = emmean, y = fire_statut , color = fire_statut)) +  
  geom_point(size = 4) +  
  geom_errorbar(aes(xmin = emmean - SE, xmax = emmean + SE), width = 0.2) + 
  labs(title = "Variation of mean CO2 efflux of the plot",
       x = "CO2 efflux (µmol/m2.sec)", y = "Fire status") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  scale_colour_manual(values = c("Burned" = "red", "Unburned" = "black"))

```


Burned plot  :  p_value =  0.55 ==> not significant (p > 0.05)
There is not significant difference between the CO2 efflux mean rate of trees in burned plot and unburned plot.
Emmean rate :
Burned  : 1.042  µmol/m2 (0.747; 1.34) 
Unburned :  0.962  µmol/m2 (0.691 ; 1.23)

```{r GLMM fire statut sp}
mod_fire_sp <- glmmTMB(CO2_efflux ~ fire_statut * species + (1|tree_tag),
  data = stem, family = gaussian(link = "identity"))

# Checking the residuals of the models 
plot(simulateResiduals(fittedModel = mod_fire_sp))

residuals_fire_sp <- residuals(mod_fire_sp)

ggplot(data = data.frame(residuals_fire_sp), aes(x = residuals_fire_sp)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Residuals histogram", x = "Residuals values", y = "Frequency") +
  theme_minimal()

summary(mod_fire_sp)

#Test Tukey emmeans fire statut sp}
emmeans_fire_sp <- emmeans(mod_fire_sp, ~ fire_statut | species)

# bo nurned vs burned = mean(emmeans )
result_fire_sp <- as.data.frame(pairs(emmeans_fire_sp, adjust = "tukey"))
result_fire_sp <- result_fire_sp %>%
  filter(!is.na(SE) & !is.na(df) & p.value <= 0.055)
result_fire_sp

estimate_fire_sp <- summary(emmeans_fire_sp)
estimate_fire_sp <- estimate_fire_sp %>%
  filter(!is.na(emmean))
estimate_fire_sp

ggplot(estimate_fire_sp, aes(x = emmean, y = species, color = fire_statut)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = emmean - SE, xmax = emmean + SE), height = 0.2) +
  labs(x = "Estimate emmean CO2 efflux (µmol/m2.sec)", y = "Species") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10)) +
  ggtitle("Effect of Fire statut on CO2 Efflux Rate by Species") + scale_colour_manual(values = c("Burned" = "red", "Unburned" = "black"))

```

species = Curatella americana: p_value =  0.0014 ==> significant 
The means of CO2 efflux are 0.469 µmol/m2.sec (0.16315  ;  0.774) on the unburned plot and 1.122 µmol/m2.sec (0.86323    1.381) burned plot. The mean rate is higher of 0.65 µmol/m2.sec in the burned plot than the unburned plot. So for this species, the presence of fire increase the mean rate of CO2 efflux and so the stem respiration.
 
 
 species = Myrcia bella: p_value =  0.0099 ==> significant
The means of CO2 efflux are 0.613 µmol/m2.sec (0.00579 ; 1.220) on the unburned plot and 1.990 µmol/m2.sec(1.13989 ; 2.841) burned plot. The difference between us are significant. The mean rate is higher of 1.37 µmol/m2.sec in the burned plot than the unburned plot. So for this species, the presence of fire increase the mean rate of CO2 efflux and so the stem respiration.
 
 species = Vochysia haenkeana: p_value =  0.0007 ==> significant
 The means of CO2 efflux are 2.905 µmol/m2.sec (2.26862 ; 3.542) on the unburned plot and 1.682 µmol/m2.sec (1.38651 ; 1.978) burned plot. The difference between us are significant (p_value = 0.0016). The mean rate is lower of 1.22 µmol/m2.sec in the burned plot than the unburned plot. So for this species, the presence of fire decrease the mean rate of CO2 efflux and so the stem respiration.

#### Effect of fire regime

```{r GLMM fire regime, echo=FALSE}

mod_regime <- glmmTMB(CO2_efflux ~ fire_regime + (1|species) + (1|tree_tag) + (1| plot_code),
  data = stem, family = gaussian(link = "identity"))

# Checking the residuals of the models 
plot(simulateResiduals(fittedModel = mod_regime))

residuals_regime <- residuals(mod_regime)

ggplot(data = data.frame(residuals_regime), aes(x = residuals_regime)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Residuals histogram", x = "Residuals", y = "Frequency") +
  theme_minimal()

summary(mod_regime)
emmeans_regime <- emmeans(mod_regime, ~ fire_regime)
result_regime <- as.data.frame(pairs(emmeans_regime, adjust = "tukey"))
result_regime

estimate_regime <- summary(emmeans_regime)

ggplot(estimate_regime, aes(y = fire_regime, x = emmean, color = fire_regime)) +  geom_point(size = 4) +  
  geom_errorbar(aes(xmin = emmean - SE, xmax = emmean + SE), width = 0.2) +  
  labs(title = "Variation of mean CO2 efflux of the plot",
       x = "CO2 efflux (µmol/m2.sec)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = color)
```
 There is no significant difference between the mean rate CO2 efflux of the fire regimes.

```{r GLMM fire regime sp, echo=FALSE}
mod_plot <- glmmTMB(CO2_efflux ~ fire_regime * species + (1|tree_tag) + (1| plot_code),
  data = stem, family = gaussian(link = "identity"))

# Checking the residuals of the models 
plot(simulateResiduals(fittedModel = mod_plot))

residuals_plot <- residuals(mod_plot)

ggplot(data = data.frame(residuals_plot), aes(x = residuals_plot)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Residuals histogram", x = "Residuals", y = "Frequency") +
  theme_minimal()

summary(mod_plot)

#Test Tukey emmeans
emmeans_plot <- emmeans(mod_plot, ~ fire_regime | species)

result_plot <- as.data.frame(pairs(emmeans_plot, adjust = "tukey"))
result_plot <- result_plot %>%
  filter(!is.na(SE) & !is.na(df) & p.value <= 0.056)
result_plot

estimate_plot <- summary(emmeans_plot)
estimate_plot <- estimate_plot %>% filter(!is.na(SE))
write.csv(estimate_plot, "C:/Users/U051-S857/OneDrive/Documents/Fac/Cours/UM/Stage M2/Dados/Result/estimate_stem_resp.csv", row.names = FALSE)

ggplot(estimate_plot, aes(y = species, x = emmean, color = fire_regime)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = emmean - SE, xmax = emmean + SE), width = 0.2) +
  labs(x = "Species", y = "CO2 efflux (µmol/m2.sec)") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 10)) +
  ggtitle("Effect of Fire regime on CO2 Efflux Rate by Species") + scale_colour_manual(values = color)

sp_plot  <- estimate_plot %>%
  filter(species %in% c("Curatella americana", "Hancornia speciosa", "Myrcia bella", "Vochysia haenkeana"))

ggplot(sp_plot, aes(y = species, x = emmean, color = fire_regime)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = emmean - SE, xmax = emmean + SE), width = 0.2) +
  labs(x = "Species", y = "CO2 efflux (µmol/m2.sec)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(title = "Effect of Fire on CO2 Efflux Mean Rate by Species",
       x = "Estimate Emmeans CO2 efflux (µmol/m2.sec)", y = "Species", color = "Fire Regime")  +
  scale_color_manual(values = color)

```

species = Curatella americana:

Unburned vs Annual: p-value = 0.037 ==> significant (p < 0.05)
The mean rate of CO2 efflux is lower by 0.76 µmol/m².sec in the "Unburned" plot than the "Annual" plot.


species = Myrcia bella:

Unburned vs Biennial: p-value = 0.007 ==> significant (p < 0.05)
The mean rate of CO2 efflux is lower by 1.38 µmol/m².sec in the "Unburned" plot than the "Biennial" plot.

species = Vochysia haenkeana:

Unburned vs Biennial: p-value = 0.0002 ==> significant (p < 0.01)
The mean rate of CO2 efflux is higher by 1.9 µmol/m².sec in the "Unburned" plot than the "Biennial" plot.

Unburned vs Annual: p-value = 0.0263 ==> significant (p < 0.05)
The mean rate of CO2 efflux is higher by 1.11 µmol/m².sec in the "Unburned" plot than the "Annual" plot.

Biennial vs Trinnial: p-value = 0.0448 ==> significant (p < 0.05)
The mean rate of CO2 efflux is lower by 0.97 µmol/m².sec in the "Biennial" plot than the "Triennial" plot.

 
## Conclusion on fire effect :

Observing the estimate mean rate of CO2 efflux of all treees in the burned and unburned plot, we detected a significant difference between them. So the presence of fire in the plot increase significaly 20% the mean rate of CO2 efflux of the tree and so, increase the mean rate of stem respiration. When we look for the mean rate of each species in burned plot, this effect changed. We detected a significant difference between 3 species : Curatella americana, Myrcia bella and Vochysia haenkeana. For Curatella americana and Myrcia bella, the presence of fire increase significaly the mean rate of CO2 efflux (~0.64 µmol/m2.s and ~1.31 µmol/m2.s) and so the stem respiration. On the contrary, the presence of fire causes a decrease of CO2 efflux Vochysia haenkeana (~ - 1.18 µmol/m2.s).

When we look at the different fire regimes in each plot, we observe significant variations in CO2 efflux rates. The unburned plot serves as the reference for the model, and the burned plots (ESA-05, ESA-07, and ESA-09) show difference significaly on the mean CO2 efflux rates than in the unburned plot. This suggests that the fire regime increase significaly the stem respiration of ~ 40% for triennial regime, ~ 20% for biennial regime and ~ 56% for annual regime.

In terms of species, the results show that some species exhibit significant differences in CO2 efflux rates across different plots. For Curatella americana, the CO2 efflux rate is higher in the annual plot the unburned, with a significant difference of 0.76 µmol/m2.s. For Hancornia speciosa, the efflux rates in unburned plot are significantly higher than in triennial plot,  with differences of 0.71 µmol/m2.s. This suggests triennial fire regime cause a decrease of the CO2 efflux and so the stem respiration of the tree in this plot. It's the contrary for Myrcia bella which shows a significant reduction of CO2 efflux in biennial plot (~ -1.31 µmol/m2.s) than in the unburned plot. So this fire regime decrease the stem respiration rates for this species. Lastly, Vochysia haenkeana exhibits significantly higher CO2 efflux in the biennial and annual plot tahn in the unburned, with significaly differences of 1.80 µmol/m2.s and 1.11  µmol/m2.s. So the presence of fire cause a decrease of stem respiration too for this species. 

In conclusion, the effect of fire on stem respiration depend of the types of species.In globaly, we find a increase of stem respiration in presence of fire, but this effect is different between the tree. It's the case for Curatela americana and Myrcia bella, but not for Vochysia haenkeana, that's trees exposed to different type of fire regime shows us a decrease of stem respiration. 
Now we has to determinate the relation between wood and bark traits with stem respiration and determinate if fire has a influence on this relation. 
