---
title: "\"Stage M2 -  Relating patterns in woody stem respiration with bark plant
  functional traits in Cerrado\""
author: "Edouard DISTIN CARVALHO"
date: "2025-04-11"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r directory, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/U051-S857/OneDrive/Documents/Fac/Cours/UM/Stage M2/Dados/Traits")
```

# Bark traits relation on Stem respiration in no burned plot

## Data 

```{r charging and cleaning data, include=FALSE}
library(readxl)
library(dplyr)

# Import dataset
bark <- read_excel("C:/Users/U051-S857/OneDrive/Documents/Fac/Cours/UM/Stage M2/Dados/Traits/bark_traits_final.xlsx")

bark <- bark %>% 
  rename(SW_D = DBH)

# Data form 
str(bark)
bark$tree_tag <-as.numeric(bark$tree_tag) 
color_plot <- c( "ESA05" = "green3" , "ESA07" = "dodgerblue", "ESA09" = "red", "ESA04" = "black")

```

## Pearson's Correlation Test 

### Bark traits in no_burned plot (ESA-04)
```{r Pearson Test, echo=FALSE}
library(corrplot)
library(MASS)
library(ggplot2)
library(GGally)
library(gridExtra)

bark_no_burned <- bark %>% filter(plot_code == "ESA04")
bark_traits <- bark_no_burned[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC")]

# Box-cox transformation for Normality condition 
boxcox_results <- apply(bark_traits, 2, function(x) {
  result <- boxcox(x ~ 1, lambda = seq(-2, 2, by = 0.1))  
  lambda_optimal <- result$x[which.max(result$y)]  
  return(lambda_optimal)
})
boxcox_results

# Box-Cox Application 

bark_2 <- bark_traits
for(i in 1:ncol(bark_2)) {
  lambda_optimal <- boxcox_results[i]
  if(lambda_optimal == 0) {
    bark_2[, i] <- log(bark_traits[, i] + 1) 
  } else {
    bark_2[, i] <- (bark_traits[, i]^lambda_optimal - 1) / lambda_optimal  
  }
}

# Normality test : Shapiro-Wilk Test 
shapiro.test(bark_2$SW_D)
shapiro.test(bark_2$OB_T)
shapiro.test(bark_2$IB_T)
shapiro.test(bark_2$TB_T)
shapiro.test(bark_2$RB_T)
shapiro.test(bark_2$B_Den)
shapiro.test(bark_2$B_WC)

# Pearson Matrix 
cor_matrix <- cor(bark_2, method = "pearson")
```


```{r Pearson matrix }
# Pearson Matrix 
cor_matrix <- cor(bark_2, method = "pearson")

corrplot(cor_matrix, method = "circle", type = "upper", order = "hclust", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45)

```

 **Positive Correlations :**
 
Outer bark thickness and Total bark thickness : 0.95 ==> very strong positive relationship 
As the outer bark thickness increases, the total bark thickness also increases.

Relative bark thickness and Outer bark thickness : 0.68 ==> positive correlation 
Thicker outer bark is associated with greater relative bark thickness.

Bark Density and diameter at breast height: 0.42  ==> positive correlation 
Larger trees may have denser bark, though the correlation is not very strong.

 **Negative Correlations :**
Relative Bark thickness and diameter at breast height : -0.70 ==> strong negative correlation
As the SW_D increases, the relative bark thickness tends to decrease, meaning that larger trees might have thinner relative bark compared to their size.

Bark water content and Outer bark thickness : -0.65 ==> negative relationship 
Thicker outer bark tends to have lower water content.

Bark water content and Total bark thickness : -0.58 s==> negative relationship
Thicker bark tends to have lower water content.

### Incluing Wood density in no burned plot
```{r With wood density, echo=FALSE}
# List of species for wood traits 
wood <- read_excel("C:/Users/U051-S857/OneDrive/Documents/Fac/Cours/UM/Stage M2/Dados/Traits/wood_density.xlsx")
wood$tree_tag <-as.numeric(wood$tree_tag) 

all_traits <- wood %>% 
  dplyr::left_join(
    bark %>% 
      dplyr::select(plot_code, sub_plot, tree_tag, family, species, code, 
                    SW_D, OB_T, IB_T, TB_T, RB_T, 
                    OB_fresh_m, IB_fresh_m, B_fresh_m, OB_v, IB_v, B_fresh_v, 
                    OB_dried_m, IB_dried_m, B_dried_m, B_Den, B_WC),
    by = c("plot_code", "sub_plot", "tree_tag", "family", "species", "code")
  ) %>% 
  dplyr::select(plot_code, sub_plot, tree_tag, family, species, code, 
                S_fresh_v, S_dried_m, SW_Den, 
                SW_D, OB_T, IB_T, TB_T, RB_T, 
                OB_fresh_m, IB_fresh_m, B_fresh_m, OB_v, IB_v, B_fresh_v, 
                OB_dried_m, IB_dried_m, B_dried_m, B_Den, B_WC)
                

traits_no_burned <- all_traits %>% filter(plot_code == "ESA04")

traits <- traits_no_burned[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]

# Box-cox transformation for Normality condition 
boxcox_results2 <- apply(traits, 2, function(x) {
  result <- boxcox(x ~ 1, lambda = seq(-2, 2, by = 0.1))  
  lambda_optimal2 <- result$x[which.max(result$y)]  
  return(lambda_optimal2)
})

# Box-Cox Application 

traits_2 <- traits
for(i in 1:ncol(traits_2)) {
  lambda_optimal2 <- boxcox_results2[i]
  if(lambda_optimal2 == 0) {
    traits_2[, i] <- log(traits[, i] + 1) 
  } else {
    traits_2[, i] <- (traits[, i]^lambda_optimal2 - 1) / lambda_optimal2  
  }
}

# Normality test : Shapiro-Wilk Test 
shapiro.test(traits_2$SW_D)
shapiro.test(traits_2$OB_T)
shapiro.test(traits_2$IB_T)
shapiro.test(traits_2$TB_T)
shapiro.test(traits_2$RB_T)
shapiro.test(traits_2$B_Den)
shapiro.test(traits_2$B_WC)
shapiro.test(traits_2$SW_Den)
```


```{r Pearson Matrix ESA04}
# Pearson Matrix 
cor_matrix2 <- cor(traits_2, method = "pearson")

# Heatmap of Pearson matrix 
corrplot(cor_matrix2, 
         method = "circle",        
         type = "upper",          
         order = "alphabet",        
         addCoef.col = "black",    
         tl.col = "black",         
         tl.srt = 45,              
         title = "Heatmap ESA-04")
```

 **Positive Correlations:**
Outer bark thickness and Total bark thickness : 0.94 ==> very strong positive relationship
As the outer bark thickness increases, the total bark thickness also increases proportionally.
  
Relative bark thickness and Outer bark thickness : 0.74 ==> positive correlation
Thicker outer bark is associated with a thicker relative bark thickness.

Wood density and Bark density: 0.56 ==> positive relationship
Trees with higher wood density tend to have thicker bark.

**Negative Correlations:**
Relative bark thickness Stem_diameter : -0.84 ==> strong negative correlation
Larger trees (with a higher diameter) tend to have a thinner relative bark thickness. 

Outer bark thickness and Bark water content : -0.60 ==> negative correlation 
Thicker bark is generally less capable of retaining water.

Total bark thickness and Bark water content: -0.49 ==> negative relationship
Thicker bark tends to have lower water content.

Relative bark thickness and Bark water content : -0.37 ==> negative relationship 
Greater relative bark thickness is associated with lower water content in the bark.

**Weak Correlations:**
There are weak to no significant linear relationships between inner bark thickness and stem diameter, between inner bark thickness and stem density

### Burned plot 

#### Biennial plot : ESA05
```{r ESA05, include = FALSE}
traits_ESA05 <- all_traits %>% filter(plot_code == "ESA05")

traits_ESA05 <- traits_ESA05[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]


# Box-cox transformation for Normality condition 
boxcox_results_ESA05 <- apply(traits_ESA05, 2, function(x) {
  result <- boxcox(x ~ 1, lambda = seq(-2, 2, by = 0.1))  
  lambda_optimal_ESA05 <- result$x[which.max(result$y)]  
  return(lambda_optimal_ESA05)
})

# Box-Cox Application 

traits_2_ESA05 <- traits_ESA05
for(i in 1:ncol(traits_2)) {
  lambda_optimal_ESA05 <- boxcox_results_ESA05[i]
  if(lambda_optimal_ESA05 == 0) {
    traits_2_ESA05[, i] <- log(traits_ESA05[, i] + 1) 
  } else {
    traits_2_ESA05[, i] <- (traits_2_ESA05[, i]^lambda_optimal_ESA05 - 1) / lambda_optimal_ESA05  
  }
}

# Normality test : Shapiro-Wilk Test 
shapiro.test(traits_2_ESA05$SW_D)
shapiro.test(traits_2_ESA05$OB_T)
shapiro.test(traits_2_ESA05$IB_T)
shapiro.test(traits_2_ESA05$TB_T)
shapiro.test(traits_2_ESA05$RB_T)
shapiro.test(traits_2_ESA05$B_Den)
shapiro.test(traits_2_ESA05$B_WC)
shapiro.test(traits_2_ESA05$SW_Den)

# Pearson Matrix 
cor_matrix_ESA05 <- cor(traits_2_ESA05, method = "pearson")
```

 **Positive Correlations:**
Outer Bark Thickness and Total Bark Thickness : 0.95 ==> Very strong positive relationship  
As the outer bark thickness increases, the total bark thickness also increases proportionally.

Outer Bark Thickness and Relative Bark Thickness : 0.78 ==> Moderate positive relationship  
Thicker bark is associated with higher relative bark thickness, which reveal a beter investment on protection against physical stress.

Stem diameter and Bark Density : 0.38 ==> Positive correlation  
Larger trees tend to have denser bark.

Bark Density and Stem Wood Density : 0.44 ==> Moderate positive relationship  
  Trees with denser bark tend to have higher stem wood density.

**Negative Correlations:**

Stem diameter and Relative Bark Thickness: -0.65 ==>bModerate negative relationship  
Larger trees tend to have lower Relative Bark Thickness, suggesting a trade-off between size and bark toughness.

Bark Density and Bark Water Content : -0.59 ==> Moderate negative correlation 
Denser bark tends to retain less water, possibly due to the compact structure of the bark.

Bark Water Content and Stem Wood Density : -0.23 ==> Negative relationship  
Higher wood density in the stem is weakly associated with lower water content in the bark.

Relative Bark Thickness and Bark Water Content : -0.16 ==> Negative relationship  
Thicker bark (with higher resistance) might have a lower ability to store water.

**Conclusion:**
The strong negative correlation between relative bark thickness the stem diameter suggests that larger trees tend to have relatively thinner bark 
= trade-offs between bark investment and other growth strategies which might be an adaptation to accommodate faster growth rates in the trunk.
For a thicker Bark, there is a strong relationship with other traits : relative bark thickness and bark density. 
= role of trees protection againts fire 

The negative relationship between bark thickness and bark water content suggests that thicker bark may have a lower capacity to retain water 
= trade-off with water retention : impact of bark on tree survival in fire-prone environments which influence their water retention capacity.


#### Triennial plot : ESA07
```{r ESA07, include= FALSE}
traits_burned <- all_traits[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]

traits_ESA07 <- all_traits %>% filter(plot_code == "ESA07")

traits_ESA07 <- traits_ESA07[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]


# Box-cox transformation for Normality condition 
boxcox_results_ESA07 <- apply(traits_ESA07, 2, function(x) {
  result <- boxcox(x ~ 1, lambda = seq(-2, 2, by = 0.1))  
  lambda_optimal_ESA07 <- result$x[which.max(result$y)]  
  return(lambda_optimal_ESA07)
})

# Box-Cox Application 

traits_2_ESA07 <- traits_ESA07
for(i in 1:ncol(traits_2)) {
  lambda_optimal_ESA07 <- boxcox_results_ESA07[i]
  if(lambda_optimal_ESA07 == 0) {
    traits_2_ESA07[, i] <- log(traits_ESA07[, i] + 1) 
  } else {
    traits_2_ESA07[, i] <- (traits_2_ESA07[, i]^lambda_optimal_ESA07 - 1) / lambda_optimal_ESA07  
  }
}

# Normality test : Shapiro-Wilk Test 
shapiro.test(traits_2_ESA07$SW_D)
shapiro.test(traits_2_ESA07$OB_T)
shapiro.test(traits_2_ESA07$IB_T)
shapiro.test(traits_2_ESA07$TB_T)
shapiro.test(traits_2_ESA07$RB_T)
shapiro.test(traits_2_ESA07$B_Den)
shapiro.test(traits_2_ESA07$B_WC)
shapiro.test(traits_2_ESA07$SW_Den)

# Pearson Matrix 
cor_matrix_ESA07 <- cor(traits_2_ESA07, method = "pearson")
```

**Positive Correlations:**

Outer Bark Thickness and Total Bark Thickness: 0.91 ==> Very strong positive relationship

Outer Bark Thickness and Relative Bark Thickness : 0.84 ==> Moderate positive relationship
Trees with thicker outer bark have relatively thicker bark overall, which is often related to better protection against environmental factors such as fire.

Total Bark Thickness and Relative Bark Thickness : 0.81 ==> Moderate positive relationship
Thicker bark is associated with higher relative bark thickness, which reveal a better investment on protection against physical stress.

Stem Wood Density (SW_Den) and Outer Bark Thickness (OB_T): 0.16 ==> Positive relationship
Trees with higher wood density also tend to have slightly thicker outer bark, though the correlation is weak.

**Negative Correlations:**

Stem Diameter and Relative Bark Thickness : -0.59 ==> Moderate negative relationship

Larger trees (with a higher stem diameter) tend to have a thinner relative bark thickness. This trade-off suggests that as trees grow larger in diameter, their bark becomes thinner relative to their size, possibly because of the increased structural demands for trunk growth.

Bark Density and Outer Bark Thickness : -0.51 ==> Moderate negative relationship
Thicker outer bark tends to be associated with lower bark density. So the thicker bark is less dense, potentially due to a more porous structure that helps with protection against fires or mechanical damage.

Bark Water Content and Outer Bark Thickness : -0.75 ==> Strong negative relationship
Thicker outer bark tends to have lower water content. So thicker bark is more rigid and dense, reducing its capacity to absorb and retain water.

Bark Density and Bark Water Content : -0.51 ==> Moderate negative relationship
This is a typical relationship where dense bark is more compact and cannot retain water because of its structural composition.

Conclusion : 

The trade-off between bark thickness and stem growth is noticeable. Larger trees appear to have relatively thinner bark, which may be an adaptation to promote trunk growth, but this reduces the relative protection of the bark.

The negative relationships between bark thickness and water content suggest a trade-off between physical protection and water storage in the bark. Moreover, trees with denser bark tend to contain less water too.So this could be reveal a implications for tree survival in fire-prone environments, where trees with thicker and denser bark may be better protected against fires but have a reduced ability to retain water.


#### Annual plot : ESA09

```{r ESA09, include=FALSE}
traits_burned <- all_traits[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]

traits_ESA09 <- all_traits %>% filter(plot_code == "ESA09")

traits_ESA09 <- traits_ESA09[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]


# Box-cox transformation for Normality condition 
boxcox_results_ESA09 <- apply(traits_ESA09, 2, function(x) {
  result <- boxcox(x ~ 1, lambda = seq(-2, 2, by = 0.1))  
  lambda_optimal_ESA09 <- result$x[which.max(result$y)]  
  return(lambda_optimal_ESA09)
})

# Box-Cox Application 

traits_2_ESA09 <- traits_ESA09
for(i in 1:ncol(traits_2)) {
  lambda_optimal_ESA09 <- boxcox_results_ESA09[i]
  if(lambda_optimal_ESA09 == 0) {
    traits_2_ESA09[, i] <- log(traits_ESA09[, i] + 1) 
  } else {
    traits_2_ESA09[, i] <- (traits_2_ESA09[, i]^lambda_optimal_ESA09 - 1) / lambda_optimal_ESA09  
  }
}

# Normality test : Shapiro-Wilk Test 
shapiro.test(traits_2_ESA09$SW_D)
shapiro.test(traits_2_ESA09$OB_T)
shapiro.test(traits_2_ESA09$IB_T)
shapiro.test(traits_2_ESA09$TB_T)
shapiro.test(traits_2_ESA09$RB_T)
shapiro.test(traits_2_ESA09$B_Den)
shapiro.test(traits_2_ESA09$B_WC)
shapiro.test(traits_2_ESA09$SW_Den)

# Pearson Matrix 
cor_matrix_ESA09 <- cor(traits_2_ESA09, method = "pearson")
```

**Positive Correlations:**
Total Bark Thickness and Outer Bark Thickness: 0.93 ==> Very strong positive relationship

Outer Bark Thickness and Relative Bark Thickness (RB_T): 0.56 ==> Moderate positive relationship
The relation between outer bark and relative bark thickness is positive but lesser than the other plot. This suggests that, in this plot, the relation is less important than in the other plot.

Bark Density and Wood Density : 0.35 ==> Moderate positive relationship
This relationship suggests that trees investing in denser bark also tend to have more compact and structurally stronger wood, but this effect is less pronounced  than in the other plot.

**Negative Correlations:**

Relative Bark Thickness and Stem Diameter : -0.61 ==> Negative relationship


Bark Water Content and Outer Bark Thickness : -0.48 ==> Moderate negative relationship
Thicker outer bark tends to have lower water content. This suggests that as the bark becomes thicker, its ability to absorb and retain water decreases. This may reflect a trade-off where thicker bark, being more rigid and compact, is better suited for protection but less efficient at water retention.

Bark Density and Bark Water Content : -0.61 ==> Negative relationship

Stem Wood Density and Bark Water Content (B_WC): -0.40 ==> Moderate negative relationship

### Trade-off 

```{r Trade-off ESA04 }
# OBT - RBT
p1 <- ggplot(traits, aes(x = OB_T, y = RB_T)) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black") + 
  labs(title = "Relation Between OBT and RBT ESA04",
       x = "Outer bark thickness (mm)",
       y = "Relative bark thickness (%)") +
  theme_classic()

model1 <- lm(RB_T ~ OB_T, data = traits)
r2_1 <- summary(model1)$r.squared  
p1 <- p1 + annotate("text", x = 5, y = 50, label = paste("R² = ", round(r2_1, 2)), color = "black", size = 5)

# B_Den - SW_Den
p2 <- ggplot(traits, aes(x = B_Den, y = SW_Den)) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black") + 
  labs(title = "Relation Between B_Den and SW_Den ESA04",
       x = "Bark density (g/cm3)",
       y = "Wood density (g/cm3)") +
  theme_classic()

model2 <- lm(SW_Den ~ B_Den, data = traits)
r2_2 <- summary(model2)$r.squared  
p2 <- p2 + annotate("text", x = 0.35, y = 0.7, label = paste("R² = ", round(r2_2, 2)), color = "black", size = 5)

# OBT - B_WC
p3 <- ggplot(traits, aes(x = OB_T, y = B_WC)) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black") + 
  labs(title = "Relation Between OBT and BWC ESA04",
       x = "Outer bark thickness (mm)" ,
       y = "Bark Water Content (%)") +
  theme_classic()

model3 <- lm(B_WC ~ OB_T, data = traits)
r2_3 <- summary(model3)$r.squared  

p3 <- p3 + annotate("text", x = 5, y = 125, label = paste("R² = ", round(r2_3, 2)), color = "black", size = 5)

# B_Den - RBT
p4 <- ggplot(traits, aes(x = B_Den, y = RB_T)) + 
  geom_point() +
  geom_smooth(method = "lm", color = "black") + 
  labs(title = "Relation Between B_Den and RBT ESA04",
       x = "Bark density (g/cm3)",
       y = "Relative Bark Thickness (%)") +
  theme_classic()

model4 <- lm(RB_T ~ B_Den, data = traits)
r2_4 <- summary(model4)$r.squared 

p4 <- p4 + annotate("text", x = 0.4, y = 25, label = paste("R² = ", round(r2_4, 2)), color = "black", size = 5)

grid.arrange(p1, p2, p3, p4, ncol = 2)

# OBT vs RBT
p5 <- ggplot(traits_ESA05, aes(x = OB_T, y = RB_T)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "green3") + 
  labs(title = "Relation Between OBT and RBT ESA05",
       x = "Outer bark thickness (mm)",
       y = "Relative bark thickness (%)") +
  theme_classic()

model5 <- lm(RB_T ~ OB_T, data = traits_ESA05)
r2_5 <- summary(model5)$r.squared  # Extraire le R²

p5 <- p5 + annotate("text", x = 5, y = 30, label = paste("R² = ", round(r2_5, 2)), color = "green3", size = 5)

# B_Den vs SW_Den
p6 <- ggplot(traits_ESA05, aes(x = B_Den, y = SW_Den)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "green3") + 
  labs(title = "Relation Between B_Den and SW_Den ESA05",
       x = "Bark density (g/cm3)",
       y = "Wood density (g/cm3)") +
  theme_classic()

model6 <- lm(SW_Den ~ B_Den, data = traits_ESA05)
r2_6 <- summary(model6)$r.squared  # Extraire le R²

p6 <- p6 + annotate("text", x = 0.2, y = 0.7, label = paste("R² = ", round(r2_6, 2)), color = "green3", size = 5)

# OBT vs B_WC
p7 <- ggplot(traits_ESA05, aes(x = OB_T, y = B_WC)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "green3") + 
  labs(title = "Relation Between OBT and B_wc ESA05",
       x = "Outer bark thickness (mm)",
       y = "Bark water content (%)") +
  theme_classic()

model7 <- lm(B_WC ~ OB_T, data = traits_ESA05)
r2_7 <- summary(model7)$r.squared  # Extraire le R²

p7 <- p7 + annotate("text", x = 20, y = 200, label = paste("R² = ", round(r2_7, 2)), color = "green3", size = 5)

# B_Den vs RBT
p8 <- ggplot(traits_ESA05, aes(x = B_Den, y = RB_T)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "green3") + 
  labs(title = "Relation Between B_Den and RBT ESA05",
       x = "Bark Density (g/cm3)",
       y = "Relative Bark Thickness (%)") +
  theme_classic()

model8 <- lm(RB_T ~ B_Den, data = traits_ESA05)
r2_8 <- summary(model8)$r.squared  # Extraire le R²

p8 <- p8 + annotate("text", x = 0.6, y = 25, label = paste("R² = ", round(r2_8, 2)), color = "green3", size = 5)


grid.arrange(p5, p6, p7, p8, ncol = 2)


# OBT vs RBT
p9 <- ggplot(traits_ESA07, aes(x = OB_T, y = RB_T)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "dodgerblue") + 
  labs(title = "Relation Between OBT and RBT ESA07",
       x = "Outer bark thickness (mm)",
       y = "Relative bark thickness (%)") +
  theme_classic()

model9 <- lm(RB_T ~ OB_T, data = traits_ESA07)
r2_9 <- summary(model9)$r.squared  # Extraire le R²

p9 <- p9 + annotate("text", x = 9, y = 30, label = paste("R² = ", round(r2_9, 2)), color = "dodgerblue", size = 5)

# B_Den vs SW_Den
p10 <- ggplot(traits_ESA07, aes(x = B_Den, y = SW_Den)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "dodgerblue") + 
  labs(title = "Relation Between B_Den and SW_Den ESA07",
       x = "Bark density (g/cm3)",
       y = "Wood density (g/cm3)") +
  theme_classic()

model10 <- lm(SW_Den ~ B_Den, data = traits_ESA07)
r2_10 <- summary(model10)$r.squared  # Extraire le R²

p10 <- p10 + annotate("text", x = 0.3, y = 0.7, label = paste("R² = ", round(r2_10, 2)), color = "dodgerblue", size = 5)

# OBT vs B_WC
p11 <- ggplot(traits_ESA07, aes(x = OB_T, y = B_WC)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "dodgerblue") + 
  labs(title = "Relation Between OBT and B_wc ESA07",
       x = "Outer bark thickness (mm)",
       y = "Bark water content (%)") +
  theme_classic()

model11 <- lm(B_WC ~ OB_T, data = traits_ESA07)
r2_11 <- summary(model11)$r.squared  # Extraire le R²

p11 <- p11 + annotate("text", x = 10, y = 200, label = paste("R² = ", round(r2_11, 2)), color = "dodgerblue", size = 5)

# B_Den vs RBT
p12 <- ggplot(traits_ESA07, aes(x = B_Den, y = RB_T)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "dodgerblue") + 
  labs(title = "Relation Between B_Den and RBT ESA07",
       x = "Bark Density (g/cm3)",
       y = "Relative Bark Thickness (%)") +
  theme_classic()

model12 <- lm(RB_T ~ B_Den, data = traits_ESA07)
r2_12 <- summary(model12)$r.squared  # Extraire le R²

p12 <- p12 + annotate("text", x = 0.4, y = 30, label = paste("R² = ", round(r2_12, 2)), color = "dodgerblue", size = 5)


grid.arrange(p9, p10, p11, p12, ncol = 2)


# OBT vs RBT
p13 <- ggplot(traits_ESA09, aes(x = OB_T, y = RB_T)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Relation Between OBT and RBT ESA09",
       x = "Outer bark thickness (mm)",
       y = "Relative bark thickness (%)") +
  theme_classic()

model13 <- lm(RB_T ~ OB_T, data = traits_ESA09)
r2_13 <- summary(model13)$r.squared  # Extraire le R²

p13 <- p13 + annotate("text", x = 5, y = 20, label = paste("R² = ", round(r2_13, 2)), color = "red", size = 5)

# B_Den vs SW_Den
p14 <- ggplot(traits_ESA09, aes(x = B_Den, y = SW_Den)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Relation Between B_Den and SW_Den ESA09",
       x = "Bark density (g/cm3)",
       y = "Wood density (g/cm3)") +
  theme_classic()

model14 <- lm(SW_Den ~ B_Den, data = traits_ESA09)
r2_14 <- summary(model14)$r.squared  # Extraire le R²

p14 <- p14 + annotate("text", x = 0.3, y = 0.7, label = paste("R² = ", round(r2_14, 2)), color = "red", size = 5)

# OBT vs B_WC
p15 <- ggplot(traits_ESA09, aes(x = OB_T, y = B_WC)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Relation Between OBT and B_wc ESA09",
       x = "Outer bark thickness (mm)",
       y = "Bark water content (%)") +
  theme_classic()

model15 <- lm(B_WC ~ OB_T, data = traits_ESA09)
r2_15 <- summary(model15)$r.squared  # Extraire le R²

p15 <- p15 + annotate("text", x = 7, y = 200, label = paste("R² = ", round(r2_15, 2)), color = "red", size = 5)

# B_Den vs RBT
p16 <- ggplot(traits_ESA09, aes(x = B_Den, y = RB_T)) + 
  geom_point(color = "black") +
  geom_smooth(method = "lm", color = "red") + 
  labs(title = "Relation Between B_Den and RBT ESA09",
       x = "Bark Density (g/cm3)",
       y = "Relative Bark Thickness (%)") +
  theme_classic()

model16 <- lm(RB_T ~ B_Den, data = traits_ESA09)
r2_16 <- summary(model16)$r.squared  # Extraire le R²

p16 <- p16 + annotate("text", x = 0.35, y = 25, label = paste("R² = ", round(r2_16, 2)), color = "red", size = 5)


grid.arrange(p13, p14, p15, p16, ncol = 2)

```

### Heatmap 
```{r Heatmap }

par(mfrow = c(2, 2))

# Heatmap of Pearson matrix 
corrplot(cor_matrix2, 
         method = "circle",        
         type = "upper",          
         order = "alphabet",        
         addCoef.col = "black",    
         tl.col = "black",         
         tl.srt = 45,              
         title = "Heatmap ESA-04")

corrplot(cor_matrix_ESA05, 
         method = "circle",        
         type = "upper",          
         order = "alphabet",        
         addCoef.col = "black",    
         tl.col = "green3",         
         tl.srt = 45,              
         title = "Heatmap ESA-05")

corrplot(cor_matrix_ESA07, 
         method = "circle", 
         type = "upper", 
         order = "alphabet",
         addCoef.col = "black", 
         tl.col = "dodgerblue", 
         tl.srt = 45, 
         title = "Heatmap ESA-07")

corrplot(cor_matrix_ESA09, 
         method = "circle", 
         type = "upper", 
         order = "alphabet",
         addCoef.col = "black", 
         tl.col = "red", 
         tl.srt = 45, 
         title = "Heatmap ESA-09")
```


**Conclusion :** 

Trade-off between growth and protection: The negative correlation between relative bark thickness (RB_T) and stem diameter (DBH) shows that as trees grow larger, they tend to allocate fewer resources to relative bark thickness. This is likely a growth strategy where larger trees prioritize trunk expansion over the investment in bark.

Bark and fire resilience: The positive relationships between bark thickness and relative bark thickness highlight the importance of bark investment in tree survival, especially in fire-prone areas. However, this comes at the cost of water retention, as seen in the negative correlation between bark thickness and water content.

Wood and bark density: The positive correlation between bark density and wood density suggests a broader structural strategy where trees with denser wood also have denser bark, potentially offering better protection against both physical damage and fire.

Water retention trade-off: Trees with thicker bark or denser bark tend to have lower water retention, indicating a trade-off between structural protection and the ability to store water, which may influence the tree's survival under drought conditions or fire stress.


## ACP 

### No burned plot : ESA04

```{r  ACP ESA04}
library(ade4)
library(ggplot2)
library(ggrepel)
library(dplyr)

esa04 <- all_traits %>% filter(plot_code == "ESA04")
data_acp04 <- esa04[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]

acp04 <- dudi.pca(data_acp04, scale = TRUE, center = TRUE, scannf = FALSE, nf = 3)

# Extraire et renommer les coordonnées des variables
var_coords <- as.data.frame(acp04$co[, 1:2])
colnames(var_coords) <- c("PC1", "PC2")
var_coords$Variable <- rownames(var_coords)

# Extraire et renommer les coordonnées des individus, puis ajouter la variable "species"
ind_coords04 <- as.data.frame(acp04$li[, 1:2])
colnames(ind_coords04) <- c("PC1", "PC2")
ind_coords04$species <- esa04$species

# Visualisation de l'acp04 avec ggplot2
plot_acp04 <- ggplot() +
  # Points pour les individus, colorés par espèce
  geom_point(data = ind_coords04, aes(x = PC1, y = PC2, color = species), alpha = 0.7) +
  # Flèches pour les variables (vecteurs)
  geom_segment(data = var_coords, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 1) +
  # Labels pour les variables
  geom_text_repel(data = var_coords, aes(x = PC1, y = PC2, label = Variable),
                  color = "black", size = 5) +
  # Axes avec pourcentages d'inertie expliquée
  xlab(paste0("PC1 (", round(acp04$eig[1] / sum(acp04$eig) * 100, 1), "%)")) +
  ylab(paste0("PC2 (", round(acp04$eig[2] / sum(acp04$eig) * 100, 1), "%)")) +
  labs(title = "acp04 - Variation selon les espèces", color = "Species") +
  theme_classic() +
  theme(legend.position = "right")

print(acp04$eig / sum(acp04$eig) * 100)  
summary(acp04)

```

### Burned plot 

#### Biennial plot : ESA05

```{r  ACP ESA05}
esa05 <- all_traits %>% filter(plot_code == "ESA05")
data_acp05 <- esa05[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]

acp05 <- dudi.pca(data_acp05, scale = TRUE, center = TRUE, scannf = FALSE, nf = 3)

# Extraire et renommer les coordonnées des variables
var_coords <- as.data.frame(acp05$co[, 1:2])
colnames(var_coords) <- c("PC1", "PC2")
var_coords$Variable <- rownames(var_coords)

# Extraire et renommer les coordonnées des individus, puis ajouter la variable "species"
ind_coords05 <- as.data.frame(acp05$li[, 1:2])
colnames(ind_coords05) <- c("PC1", "PC2")
ind_coords05$species <- esa05$species

# Visualisation de l'acp05 avec ggplot2
plot_acp05 <- ggplot() +
  # Points pour les individus, colorés par espèce
  geom_point(data = ind_coords05, aes(x = PC1, y = PC2, color = species), alpha = 0.7) +
  # Flèches pour les variables (vecteurs)
  geom_segment(data = var_coords, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 1) +
  # Labels pour les variables
  geom_text_repel(data = var_coords, aes(x = PC1, y = PC2, label = Variable),
                  color = "black", size = 5) +
  # Axes avec pourcentages d'inertie expliquée
  xlab(paste0("PC1 (", round(acp05$eig[1] / sum(acp05$eig) * 100, 1), "%)")) +
  ylab(paste0("PC2 (", round(acp05$eig[2] / sum(acp05$eig) * 100, 1), "%)")) +
  labs(title = "acp05 - Variation selon les espèces", color = "Species") +
  theme_classic() +
  theme(legend.position = "right")

print(acp05$eig / sum(acp05$eig) * 100)  
summary(acp05)

```

#### Triennial plot : ESA07

```{r  ACP ESA07}
esa07 <- all_traits %>% filter(plot_code == "ESA07")
data_acp07 <- esa07[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]

acp07 <- dudi.pca(data_acp07, scale = TRUE, center = TRUE, scannf = FALSE, nf = 3)

# Extraire et renommer les coordonnées des variables
var_coords <- as.data.frame(acp07$co[, 1:2])
colnames(var_coords) <- c("PC1", "PC2")
var_coords$Variable <- rownames(var_coords)

# Extraire et renommer les coordonnées des individus, puis ajouter la variable "species"
ind_coords07 <- as.data.frame(acp07$li[, 1:2])
colnames(ind_coords07) <- c("PC1", "PC2")
ind_coords07$species <- esa07$species

# Visualisation de l'acp07 avec ggplot2
plot_acp07 <- ggplot() +
  # Points pour les individus, colorés par espèce
  geom_point(data = ind_coords07, aes(x = PC1, y = PC2, color = species), alpha = 0.7) +
  # Flèches pour les variables (vecteurs)
  geom_segment(data = var_coords, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 1) +
  # Labels pour les variables
  geom_text_repel(data = var_coords, aes(x = PC1, y = PC2, label = Variable),
                  color = "black", size = 5) +
  # Axes avec pourcentages d'inertie expliquée
  xlab(paste0("PC1 (", round(acp07$eig[1] / sum(acp07$eig) * 100, 1), "%)")) +
  ylab(paste0("PC2 (", round(acp07$eig[2] / sum(acp07$eig) * 100, 1), "%)")) +
  labs(title = "acp07 - Variation selon les espèces", color = "Species") +
  theme_classic() +
  theme(legend.position = "right")

print(acp07$eig / sum(acp07$eig) * 100)  
summary(acp07)

```

#### Annual plot : ESA09

```{r ACP ESA09}
esa09 <- all_traits %>% filter(plot_code == "ESA09")
data_acp09 <- esa09[, c("SW_D","OB_T", "IB_T", "TB_T", "RB_T", "B_Den", "B_WC", "SW_Den")]

acp09 <- dudi.pca(data_acp09, scale = TRUE, center = TRUE, scannf = FALSE, nf = 3)

# Extraire et renommer les coordonnées des variables
var_coords <- as.data.frame(acp09$co[, 1:2])
colnames(var_coords) <- c("PC1", "PC2")
var_coords$Variable <- rownames(var_coords)

# Extraire et renommer les coordonnées des individus, puis ajouter la variable "species"
ind_coords09 <- as.data.frame(acp09$li[, 1:2])
colnames(ind_coords09) <- c("PC1", "PC2")
ind_coords09$species <- esa09$species

# Visualisation de l'acp09 avec ggplot2
plot_acp09 <- ggplot() +
  # Points pour les individus, colorés par espèce
  geom_point(data = ind_coords09, aes(x = PC1, y = PC2, color = species), alpha = 0.7) +
  # Flèches pour les variables (vecteurs)
  geom_segment(data = var_coords, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 1) +
  # Labels pour les variables
  geom_text_repel(data = var_coords, aes(x = PC1, y = PC2, label = Variable),
                  color = "black", size = 5) +
  # Axes avec pourcentages d'inertie expliquée
  xlab(paste0("PC1 (", round(acp09$eig[1] / sum(acp09$eig) * 100, 1), "%)")) +
  ylab(paste0("PC2 (", round(acp09$eig[2] / sum(acp09$eig) * 100, 1), "%)")) +
  labs(title = "acp09 - Variation selon les espèces", color = "Species") +
  theme_classic() +
  theme(legend.position = "right")

print(acp09$eig / sum(acp09$eig) * 100)  
summary(acp09)

```

### Correlation circle 

```{r ACP Circle}
par(mfrow = c(2,2))

s.corcircle(acp04$co, sub = "ACP Circle ESA04")
s.corcircle(acp05$co, sub = "ACP Circle ESA05")
s.corcircle(acp07$co, sub = "ACP Circle ESA07")
s.corcircle(acp09$co, sub = "ACP Circle ESA09")

plot_acp04
plot_acp05
plot_acp07
plot_acp09

```

